---
title: "Part 2"
author: "Aishameriane Schmidt"
date: "5 de maio de 2017"
output: html_document
---

## Part 2 - Vectors

Vamos associar "Bora!" à variável "Vegas":

```{r}
Vegas<-"Bora!"
```

### Create a vector (2)

O R utiliza a função `c()` (de concatenar) para criação de vetores. Os valores são dispostos dentro do parênteses, separados por vírgulas:

```{r}
numeric_vector <- c(1, 10, 49)
character_vector <- c("a", "b", "c")
boolean_vector <- c(T,F,T)
```

### Create a vector (3)

After one week in Las Vegas and still zero Ferraris in your garage, you decide that it is time to start using your data analytical superpowers.

Before doing a first analysis, you decide to first collect all the winnings and losses for the last week:

For poker_vector:

* On Monday you won $140
* Tuesday you lost $50
* Wednesday you won $20
* Thursday you lost $120
* Friday you won $240

For roulette_vector:

* On Monday you lost $24
* Tuesday you lost $50
* Wednesday you won $100
* Thursday you lost $350
* Friday you won $10

You only played poker and roulette, since there was a delegation of mediums that occupied the craps tables. To be able to use this data in R, you decide to create the variables poker_vector and roulette_vector.

```{r}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <-  c(-24, -50, 100, -350, 10)
```

### Naming a vector (1)

s a data analyst, it is important to have a clear view on the data that you are using. Understanding what each element refers to is therefore essential.

In the previous exercise, we created a vector with your winnings over the week. Each vector element refers to a day of the week but it is hard to tell which element belongs to which day. It would be nice if you could show that in the vector itself.

You can give a name to the elements of a vector with the names() function. Have a look at this example:

some_vector <- c("John Doe", "poker player")
names(some_vector) <- c("Name", "Profession")
This code first creates a vector some_vector and then gives the two elements a name. The first element is assigned the name Name, while the second element is labeled Profession. Printing the contents to the console yields following output:
`Name`       `Profession` 
`"John Doe"` `"poker player"` 

```{r}
# Assign days as names of poker_vector
names(poker_vector) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

# Assign days as names of roulette_vectors
names(roulette_vector)<- names(poker_vector)
```

### Naming a vector (2)

If you want to become a good statistician, you have to become lazy. (If you are already lazy, chances are high you are one of those exceptional, natural-born statistical talents.)

In the previous exercises you probably experienced that it is boring and frustrating to type and retype information such as the days of the week. However, when you look at it from a higher perspective, there is a more efficient way to do this, namely, to assign the days of the week vector to a variable!

Just like you did with your poker and roulette returns, you can also create a variable that contains the days of the week. This way you can use and re-use it.

```{r}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# The variable days_vector
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
 
# Assign the names of the day to roulette_vector and poker_vector
names(poker_vector) <-   days_vector
names(roulette_vector) <- days_vector
```

### Calculating total winnings (1)
Now, to compute some poker and roulette metrics, you have to do arithmetic calculations on vectors.

It is important to know that if you sum two vectors in R, it takes the element-wise sum. For example, the following three statements are completely equivalent:

```{r}
c(1, 2, 3) + c(4, 5, 6)
c(1 + 4, 2 + 5, 3 + 6)
c(5, 7, 9)
```

You can also do the calculations with variables that represent vectors:

```{r}
a <- c(1, 2, 3) 
b <- c(4, 5, 6)
c <- a + b
```

See example below: two vectors, A and B are added into a new one, named `total_vector`.

```{r}
A_vector <- c(1, 2, 3)
B_vector <- c(4, 5, 6)

# Take the sum of A_vector and B_vector
total_vector <- A_vector + B_vector
  
# Print out total_vector
total_vector
```


### Calculating total winnings (2)

Now you understand how R does arithmetic with vectors, it is time to get those Ferraris in your garage! First, you need to understand what the overall profit or loss per day of the week was. The total daily profit is the sum of the profit/loss you realized on poker per day, and the profit/loss you realized on roulette per day.

In R, this is just the sum of roulette_vector and poker_vector.

```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Assign to total_daily how much you won/lost on each day
total_daily <- poker_vector + roulette_vector
```

### Calculating total winnings (3)

Based on the previous analysis, it looks like you had a mix of good and bad days. This is not what your ego expected, and you wonder if there may be a very tiny chance you have lost money over the week in total?

A function that helps you to answer this question is sum(). It calculates the sum of all elements of a vector. For example, to calculate the total amount of money you have lost/won with poker you do:

```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Total winnings with poker
total_poker <- sum(poker_vector)

# Total winnings with roulette
total_roulette <-  sum(roulette_vector)

# Total winnings overall
total_week <- total_poker + total_roulette

# Print out total_week
  total_week
```

### Comparing total winnings

Parece que a sorte não está do seu lado. Vamos tentar analisar melhor como que foi o seu desempenho. Talvez suas habilidades na roleta não sejam tão boas quanto as suas habilidades em pôker e por isso seus ganhos médios estão negativos.

```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Calculate total gains for poker and roulette
total_poker <- sum(poker_vector)
total_roulette <- sum(roulette_vector)

# Check if you realized higher total gains in poker than in roulette 
total_poker > total_roulette
```

### Vector selection: the good times (1)

Aparentemente seu palpite estava correto: os ganhos com jogos de pôker foram em média maiores que os de roleta. Porém, é possível que este não seja um comportamento sistemático e outra análise interessante é comparar seu desempenho ao longo dos dias da semana. Para acessar posições específicas de um vetor, é possível utilizar colchetes `[]`, como por exemplo `total_poker[1]` irá exibir os ganhos da segunda feira. Vamos salvar os ganhos da quarta-feira no jogo de poker:

```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Define a new variable based on a selection
poker_wednesday <-poker_vector[3]
poker_wednesday

# Outra possibilidade
poker_wednesday <- poker_vector["Wednesday"]
poker_wednesday
```

### Vector selection: the good times (2)

Para selecionar múltiplos elementos de um vetor, você pode, por exemplo, indicar dentro dos colchetes quais elementos quer, utilizando a função concatenar. O comando para acessar o primeiro e o quinto elemento do vetor de ganhos do poker seria `r poker_vector[c(1,5)]`.

```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Define a new variable based on a selection
poker_midweek<-poker_vector[c(2,3,4)]

# Outra possibilidade
poker_midweek <- poker_vector[2:4]
```

### Vector selection: the good times (4)

Podemos selecionar os pedaços de vetores e armazenar em um outro vetor para a partir daí criar métricas. Por exemplo, para saber os ganhos do poker de segunda a quarta feira, podemos utilizar o seguinte código:

```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Select poker results for Monday, Tuesday and Wednesday
poker_start <- poker_vector[1:3]
  
# Calculate the average of the elements in poker_start
mean(poker_start)
```

### Selection by comparison - step 1

Podemos fazer comparações entre elementos de vetores ou entre vetores e números. Os comparadores lógicos disponíveis em R são:

* `<` para "menor que";
* `>` para "maior que";
* `<=` para "menor ou igual que";
* `>=` para "maior ou igual que";
* `==` para "igual que";
* `!=` para "não igual a".

Conforme visto nas lições anteriores, se digitarmos no console `r 6>5` o resultado será `TRUE`. Podemos utilizar esse mesmo método para vetores:

```{r}
c(4,5,6) > 5
```

Vamos verificar em quais dias os ganhos no poker foram positivos:

```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on poker?
selection_vector <- poker_vector > 0
  
# Print out selection_vector
selection_vector
```

### Selection by comparison - step 2

Se agora quisermos saber não só em quais dias nós tivemos ganhos positivos mas também qual o valor dos ganhos, podemos usar a sintaxe `poker_vector[selection_vector]`. Lembre-se que `selection_vector` é um vetor lógico. Ao usar o comando `poker_vector[selection_vector]` estamos pedindo ao R que exiba somente as entradas em `poker_vector` correspondentes a `selection_vector == 1`.

```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on poker?
selection_vector <- poker_vector > 0

# Select from poker_vector these days
poker_winning_days <- poker_vector[selection_vector]
poker_winning_days
```

### Advanced selection

Vamos fazer o mesmo procedimento para os jogos de roleta:

```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on roulette?
selection_vector <- roulette_vector > 0

# Select from roulette_vector these days
roulette_winning_days <- roulette_vector[selection_vector]
```

